---
title: "Sahadeo_Assignment5-3"
author: "Rishi Sahadeo"
date: "2025-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-data}
# load data set
brewing <- read.table("C:/Users/supaq/OneDrive/Desktop/Rutgers Courses/Regression Methods - Fall 2025/brewing_data.txt", header = TRUE, sep = "\t")

dim(brewing)
head(brewing, 5)

```

### 1) Visually display the data in an appropriate graph and comment on anything that may be of note. In particular, are the assumptions needed for fitting the simple linear model met?

```{r plot-brewing}
plot(brewing$Temperature, brewing$Alcohol_Percentage,
     xlab = "Temperature (F)",
     ylab = "Alcohol Percentage",
     main = "Alcohol vs Temperature",
     pch = 19)

```

The scatterplot of Alcohol Percentage vs. Temperature shows an overall negative linear trend. As temperature increases, alcohol percentage tends to decrease. One point is a clear outlier with a negative alcohol percentage which must be addressed. The remaining data appear to cluster fairly evenly around a straight line trend. Whether the variance is constant cannot be concluded from the scatterplot alone.

### 2) Initially, the brewer would just like to get a rough estimate of what the alcohol content would be if he ferments the batch at a given temperature. Are enough of the regression assumptions satisfied so that simple linear regression can be used towards the prior-mentioned goal? If not, what deviations do you need to address and how do you address them? 

```{r clean-flag}
which(brewing$Alcohol_Percentage < 0)
```

```{r clean-data}
brewing_clean <- subset(brewing, Alcohol_Percentage >= 0)
dim(brewing);
dim(brewing_clean)
```

For the brewer's goal of obtaining a rough estimate of alcohol content at a given temperature, the assumptions of a simple linear regression are reasonably well met after addressing one clear violation. The scatterplot showed a negative alcohol percentage value at observation 186. Since alcohol content cannot be negative, this is a data entry or measurement error and was excluded from the analysis.

```{r plot-clean}
plot(brewing_clean$Temperature, brewing_clean$Alcohol_Percentage,
     xlab = "Temperature (°F)",
    ylab = "Alcohol (%)",
     main = "Alcohol vs Temperature — After Removing Outlier",
     pch = 19)

```

After removing the negative alcohol value, the scatterplot shows a clear negative linear relationship. As temperature increases, alcohol percentage decreases. The points cluster closely around a straight-line trend, with no obvious curvature or extreme deviations.

### 3) After addressing any necessary issues in part 2, fit the simple linear model to the data. Provide the parameter estimates and the R2 value.

```{r fit-slr}
brew_fit <- lm(Alcohol_Percentage ~ Temperature, data = brewing_clean)
brew_fit

```

```{r fit-summary}
summary(brew_fit)

```

```{r fit-extract}
coef(brew_fit)
summary(brew_fit)$r.squared

```

Fitting a simple linear regression of Alcohol % on Temperature (F) using the clean data yields, Alcohol% = 26.174 - 0.358 * Temperature. 
A 1 degree F increase in temperature is associated with an average 0.358 percentage-point decrease in alcohol content within the observed temperature range. The intercept 26.174% is the fitted mean at 0 degree F. The models R^2 = 0.869 indicates that about 86.9% of the variability in alcohol percentage is explained by temperature in this linear model. The estimates and R^2 support a strong negative linear relationship in this dataset.

Estimates: Intercept = 26.174, Slope = -0.358, R^2 = 0.869.




